---
output: 
  pdf_document:
    fig_caption: true
    dev: pdf
fontsize: 11pt
geometry: margin=1in
header-includes:
   - \usepackage{float}
   - \usepackage{caption}
   - \captionsetup{labelformat=empty}
   - \AtBeginDocument{\let\maketitle\relax}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = "center",
  fig.pos = 'H',
  dpi = 300,        # High resolution for any remaining raster elements
  out.width = "100%" # Ensures plots fit the page width
)

library(tidyverse)
library(kableExtra)

# Global theme from your R script
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(size = 14, face = "bold"),
      legend.position = "none"
    )
)
```
 
---
 
```{=latex}
\begin{center}
\Huge \textbf{Pipeline Preprocess Report}

\vspace{0.5cm}

\large
\end{center}
```
---

Report generated on `r format(Sys.Date(), '%B %d, %Y')`.<br>

\vspace{0.5cm}

  # Project Info
  
```{r}
project_info <- read_tsv("project_info.tsv", show_col_types = FALSE)

project_info %>%
  kable(
    booktabs = TRUE,
    format = "latex",
    longtable = FALSE
  ) %>%
  kable_styling(
    latex_options = c("hold_position","scale_down"),
    font_size = 10
  ) %>%
  column_spec(1, width = "6cm") %>%
  column_spec(2, width = "8cm")
```

---

  # Overview
  
  This report documents how the dataset fared during the pipeline preprocessing.

  Briefly, dimer content is estimated in raw files and read stats are generated across the rest of the Preprocess (rainbow_bridge steps in this order):
  
  - trim_merge  
  - ngsfilter  
  - length_filtered  
  - relabeled  
  - merged  

---

\newpage

  # Helper Scripts
  
```{r}
helper_data <- data.frame(
  `Helper Script` = c("check_dimer.sh", "run_preprocess_summaries.sbatch", "plot_preprocess_results.R"),
  Process = c("Estimates dimer content across raw files", 
              "Generates read statistic before and after each step", 
              "Makes plots for the preprocessing report")
)

helper_data %>%
  kable(booktabs = TRUE, format = "latex", col.names = c("Helper Script", "Process")) %>%
  kable_styling(latex_options = "hold_position") %>%
  column_spec(1, italic = TRUE, color = "blue")
``` 
  
---
  
  # Estimated Dimer Content Across Files
```{r}
dimer_data <- read_tsv("average_primer_dimer.tsv", show_col_types = FALSE)

dimer_data %>%
  kable(
    booktabs = TRUE,
    format = "latex",
    caption = "PreTable 1: Average read and primer dimer content across input files."
  ) %>%
  kable_styling(latex_options = c("hold_position", "striped"))
```

  ---

\newpage

# Read abundance, loss, length, and alignment in dataset

```{r, fig.cap="PreFig 1: Number of reads across samples after each preprocess step"}
readcount <- read_tsv("summary-readcount_preprocess.tsv", show_col_types = FALSE)

readcount <- readcount %>%
  mutate(raw = rowMeans(select(., raw_F, raw_R))) %>%
  select(sample, raw, trim_merge, ngsfilter, l_filtered, relabeled)

readcount_long <- readcount %>%
  pivot_longer(-sample, names_to = "stage", values_to = "reads")

readcount_long$stage <- factor(
  readcount_long$stage,
  levels = c("raw", "trim_merge", "ngsfilter", "l_filtered", "relabeled")
)

stage_totals <- readcount_long %>%
  group_by(stage) %>%
  summarise(total_reads = sum(reads, na.rm = TRUE))

pre1 <- ggplot(stage_totals, aes(stage, total_reads, fill = stage)) +
  geom_col() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Total Reads per Preprocessing Stage", x = "Stage", y = "Total Reads")

pre1
```

\newpage

```{r, fig.cap="PreFig 2: Distribution of reads across samples after each preprocess step."}
pre2 <- ggplot(readcount_long, aes(stage, reads, fill = stage)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Distribution of Read Counts by Preprocessing Stage", x = "Stage", y = "Read Count")

pre2
```
  
\newpage
 
```{r, fig.cap="PreFig 3: Number of reads lost across samples after each preprocess step."}
readcount_loss_abs <- readcount %>%
  mutate(
    loss_trim_merge = raw - trim_merge,
    loss_ngsfilter = trim_merge - ngsfilter,
    loss_l_filtered = ngsfilter - l_filtered,
    loss_relabeled = l_filtered - relabeled
  ) %>%
  pivot_longer(starts_with("loss_"), names_to = "stage", values_to = "read_loss") %>%
  mutate(stage = factor(str_remove(stage, "loss_"), 
                        levels = c("trim_merge", "ngsfilter", "l_filtered", "relabeled")))

stage_loss_totals <- readcount_loss_abs %>%
  group_by(stage) %>%
  summarise(total_read_loss = sum(read_loss, na.rm = TRUE))

pre3 <- ggplot(stage_loss_totals, aes(stage, total_read_loss, fill = stage)) +
  geom_col() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Total Read Loss per Preprocessing Stage", x = "Stage", y = "Total Read Loss")

total_raw_reads <- sum(readcount$raw, na.rm = TRUE)
stage_loss_percent <- stage_loss_totals %>%
  mutate(percent_total_loss = total_read_loss / total_raw_reads * 100)

pre3
```

\newpage

```{r, fig.cap="PreFig 4: Percent read loss across samples after each preprocess step."}
pre4 <- ggplot(stage_loss_percent, aes(stage, percent_total_loss, fill = stage)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(title = "Percent Total Read Loss per Preprocessing Stage", x = "Stage", y = "Percent Loss")

pre4
```

\newpage

```{r, fig.cap="PreFig 5: Average read length across samples after each preprocess step."}
readL_step <- read_tsv("summary-readL_preprocess.tsv", show_col_types = FALSE)
step_levels <- c("raw", "trim_merge", "ngsfilter", "length_filtered", "relabeled", "merged")

ave_readL <- readL_step %>%
  group_by(step) %>%
  summarise(ave_read_l = round(mean(length, na.rm = TRUE), 0)) %>%
  mutate(step = factor(step, levels = step_levels))

pre5 <- ggplot(ave_readL, aes(step, ave_read_l, fill = step)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = ave_read_l), vjust = -0.5, size = 3.5) +
  labs(title = "Average Read Length Across Preprocessing Steps", x = "Step", y = "Average Read Length (bp)") +
  ylim(0, max(ave_readL$ave_read_l) * 1.1)

# Alignment Logic
readalign <- read_tsv("summary-readalign_trim_merge.tsv", show_col_types = FALSE)
total_val <- readalign %>% filter(read_type == "total") %>% pull(number)
rounded_total <- round(total_val)
components <- readalign %>% 
  filter(read_type != "total") %>%
  mutate(percent = round(number / rounded_total * 100, 1))

custom_levels <- c("unaligned", "aligned_full-L_collapsed", "aligned_trunc_collapsed", "aligned_not-collapsed")
components$read_type <- factor(components$read_type, levels = custom_levels)
components <- components %>% arrange(read_type)
legend_labels <- setNames(paste0(components$read_type, " (", components$percent, "%)"), components$read_type)

pre5
```

\newpage

```{r, fig.cap="PreFig 6: Relative percent of reads per trimming category in trim\\_merge"}
pre6 <- ggplot(components, aes(x = 1, y = number, fill = read_type)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar(theta="y") +
  xlim(0.5, 1.5) +
  theme_void() +
  labs(title = "Read Alignment in trim_merge") +
  annotate("text", x=0, y=0, label=paste0("Total Reads\n", format(rounded_total, big.mark=",")), size=4.5) +
  scale_fill_discrete(labels = legend_labels) +
  theme(legend.position = "right", legend.text = element_text(size = 9), plot.title = element_text(hjust = 0.5, face="bold"))

pre6
```

\newpage
```{r}
n_stages <- length(unique(readcount_long$stage))
fig_height <- n_stages * 2.5
```
```{r, fig.cap="PreFig 7: Number of reads remaining after each preprocess step.", fig.width=10, fig.height=length(unique(readcount_long$stage)) * 2.5}
mean_reads <- readcount_long %>%
  group_by(stage) %>%
  summarise(mean_reads = mean(reads, na.rm = TRUE), 
            label = scales::comma(round(mean_reads)))

ggplot(readcount_long, aes(sample, reads)) +
  geom_col(fill = "#2c7fb8") +
  geom_hline(data = mean_reads, aes(yintercept = mean_reads), 
             linetype="dashed", color="red", linewidth=0.6) +
  geom_label(data = mean_reads, aes(x = 1, y = mean_reads, label = label), 
             color="red", size=4, fill="white", hjust=0) +
  facet_wrap(~stage, ncol = 1, scales="free_y") +
  labs(title = "Read Counts per Sample by Preprocessing Stage", x = "Sample", y = "Read Count") +
  theme(axis.text.x = element_text(angle = 90, size = 6))
```

\newpage

```{r, fig.cap="PreFig 8: Percent of read loss after each preprocess step.", fig.width=10, fig.height=length(unique(readcount_long$stage)) * 2.5} 
readcount_loss_long <- read_tsv("summary-readcount_preprocess.tsv", show_col_types = FALSE) %>%
  select(sample, starts_with("%_loss_")) %>%
  pivot_longer(-sample, names_to="stage", values_to="percent_loss") %>%
  mutate(stage = factor(str_remove(stage, "%_loss_"), 
                        levels=c("trim_merge", "ngsfilter", "l_filtered", "relabeled")))

mean_loss <- readcount_loss_long %>%
  group_by(stage) %>%
  summarise(mean_loss = mean(percent_loss, na.rm = TRUE), 
            label = paste0(round(mean_loss,1),"%"))

ggplot(readcount_loss_long, aes(sample, percent_loss)) +
  geom_col(fill="#d95f02") +
  geom_hline(data=mean_loss, aes(yintercept=mean_loss), 
             linetype="dashed", color="red", linewidth=0.6) +
  geom_label(data=mean_loss, aes(x=1, y=mean_loss, label=label), 
             color="red", size=4, fill="white", hjust=0) +
  facet_wrap(~stage, ncol=1, scales="free_y") +
  labs(title="Read Loss per Sample by Preprocessing Stage", x = "Sample", y = "Percent Read Loss") +
  theme(axis.text.x = element_text(angle = 90, size = 6))
```

\newpage

```{r, fig.cap="PreFig 9: Frequency of read lengths after each preprocess step. Dashed line shows the mean.", fig.width=10, fig.height=length(unique(readcount_long$stage)) * 2.5}
summary_readL_step <- readL_step %>%
  group_by(step) %>%
  summarise(mean_length = mean(length))

ggplot(readL_step, aes(length)) +
  geom_histogram(binwidth = 5, fill="#2c7fb8", alpha=0.7) +
  facet_wrap(~step, ncol=1, scales="free_y") +
  geom_vline(data = summary_readL_step, aes(xintercept = mean_length), 
             color="red", linetype="dashed", linewidth=0.8) +
  labs(title="Read Length Distributions per Preprocessing Step", x="Read Length (bp)", y="Count") +
  theme(strip.text = element_text(size = 11))
```
